---
title: "CSCI E-63C Week 12 Problem Set"
author: 'Joshua Sacher'
date: '`r Sys.Date()`'

output:
  html_document:
    df_print: kable
    number_sections: true
    toc: true
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
library(psych)
library(ISLR)
library(e1071)
library(randomForest)
library(class)
library(ggplot2)
library(GGally)
library(scatterplot3d)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, 
                      fig.width = 6, fig.height = 6)
```


# Preface

This week problem set will explore behavior of support vector classifiers and SVMs (following the distinction made in ISLR) on WiFi localization dataset from UCI ML archive.  We worked with it on multiple occasions before (most recently two weeks ago evaluating performance of logistic regression, discriminant analysis and KNN on it).  As two weeks ago we are going to convert the four-levels outcome in the data file to the binary one indicating localization at the third location:

```{r wifiExample,fig.width=8,fig.height=8,warning=FALSE}
wifiLocDat <- read.table("wifi_localization.txt", sep = "\t")
colnames(wifiLocDat) <- c(paste0("WiFi", 1:7), "Loc")
ggpairs(wifiLocDat, aes(colour = factor(Loc)), progress = FALSE)
wifiLocDat[, "Loc3"] <- factor(wifiLocDat[, "Loc"] == 3)
wifiLocDat <- wifiLocDat[, colnames(wifiLocDat) != "Loc"]
dim(wifiLocDat)
summary(wifiLocDat)
head(wifiLocDat)
```

Here we will use SVM implementation available in library `e1071` to fit classifiers with linear and radial (polynomial for extra points) kernels and compare their relative performance as well as to that of random forest and KNN.

## `assess.prediction` from HW 10

```{r}
assess.prediction <- function(actual, predicted, print.vals = TRUE) {
  # Convert to logical values if needed
  actual = as.logical(actual)
  predicted = as.logical(predicted)
  
  TP = sum(actual & predicted)
  TN = sum(!actual & !predicted)
  FP = sum(!actual & predicted)
  FN = sum(actual & !predicted)
  P = TP+FN # total number of positives in the actual data
  N = FP+TN # total number of negatives
  
  if (print.vals) {
    # For prettiness
    cat("\nCLASSIFIER PERFORMANCE\n")
    # Accuracy (TP + TN) / (TP + TN + FP + FN)
    cat("Accuracy:        ", round(100 * (TP + TN) / (P + N), 1), "%\n", sep ="")
    # Sensitivity TP / (TP + FN)
    cat("Sensitivity:     ", round(100 * TP / P, 1), "%\n", sep ="")
    # Specificity TN / (TN + FP)
    cat("Specificity:     ", round(100 * TN / N, 1), "%\n", sep ="")
    # Precision (Positive Predictive Value) TP / (TP + FP)
    cat("Precision:       ", round(100 * TP / (TP + FP), 1), "%\n", sep ="")
    # False discovery rate 1 - PPV
    cat("False Discovery: ", round(100 * FP / (TP + FP), 1), "%\n", sep ="")
    # False positive rate 1 - TNR
    cat("False Positive:  ", round(100 * FP / N, 1), "%\n", sep ="")
    cat("\n")
  }
  
  # Return table for printing
  return(table(actual = actual, predicted = predicted))
}
```

```{r}
# N controlls the number of replicates in all replication loops
N <- 10
```


# Problem 1 (20 points): support vector classifier (i.e. using linear kernel) 

Use `svm` from library `e1071` with `kernel="linear"` to fit classifier (e.g. ISLR Ch.9.6.1) to the entire WiFi localization dataset setting parameter `cost` to 0.001, 1, 1000 and 1 mln.  Describe how this change in parameter `cost` affects model fitting process (hint: the difficulty of the underlying optimization problem increases with cost -- can you explain what drives it?) and its outcome (how does the number of support vectors change with `cost`?) and what are the implications of that.  Explain why change in `cost` value impacts number of support vectors found. (Hint: there is an answer in ISLR.)  Use `tune` function from library `e1071` (see ISLR Ch.9.6.1 for details and examples of usage) to determine approximate value of cost (in the range between 0.1 and 100 -- the suggested range spanning ordes of magnitude should hint that the density of the grid should be approximately logarithmic -- e.g. 1, 3, 10, ... or 1, 2, 5, 10, ... etc.) that yields the lowest error in cross-validation employed by `tune`.  Setup a resampling procedure repeatedly splitting entire dataset into training and test, using training data to `tune` cost value and test dataset to estimate classification error. Report and discuss distributions of test errors from this procedure and selected values of `cost`.

### Changes in cost

### Code

```{r SVM linear kernal, cache=TRUE}
for (cost in c(1e-3, 1, 1e3, 1e6)) {
  svm.result <-
    svm(Loc3 ~ .,
        data = wifiLocDat,
        kernel = "linear",
        cost = cost)
  # summary(svm.result)
  print(paste("cost =", cost))
  assess.prediction(actual = wifiLocDat$Loc3, 
                    predicted = predict(svm.result))
  plot(
    svm.result,
    data = wifiLocDat,
    formula = WiFi5 ~ WiFi1,
    grid = 250
  )
  title(sub = paste("cost =", cost))
}
```

### Discussion

## Use of `tune()` to get best cost

### Code

```{r SVM tune for cost, cache=TRUE}
svm.acc <- numeric()
costs <- numeric()
for (i in 1:N) {
  train <- sample(c(TRUE, FALSE), size = nrow(wifiLocDat), replace = TRUE)
  res <- tune(svm, Loc3 ~ ., data = wifiLocDat[train,], kernel="linear",
              ranges = list(cost = 2 ^ (-4:7)))
  p <- assess.prediction(wifiLocDat[!train,]$Loc3, 
                         predict(res$best.model, newdata = wifiLocDat[!train,]),
                         print.vals = FALSE)
  svm.acc <- c(svm.acc, sum(diag(p)) / sum(p))
  costs <- c(costs, res$best.parameters$cost)
}
```

```{r SVM tune plots and analysis}
temp <- data.frame(measure = c(rep("accuracy", length(svm.acc)),
                               rep("cost", length(costs))),
                   value = c(svm.acc, costs))
ggplot(data = temp, aes(x = "parameter", y = value, fill = measure)) + 
  facet_wrap(temp$measure, scales = "free") + 
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_jitter(height = 0)

plot(costs, svm.acc)
plot(res)

describe(svm.acc)
describe(costs)
```

### Discussion

# Problem 2 (10 points): comparison to random forest

Fit random forest classifier on the entire WiFi localization dataset with default parameters.  Calculate resulting misclassification error as reported by the confusion matrix in random forest output.  Explain why error reported in random forest confusion matrix represents estimated test (as opposed to train) error of the procedure.  Compare resulting test error to that for support vector classifier obtained above and discuss results of such comparison.

## Code

```{r RF classifier, cache=TRUE}
# Using default settings (just passing in formula and data)
wifi.rf <- randomForest(Loc3 ~ ., data = wifiLocDat)

# Confusion matrix
wifi.rf$confusion

# Accuracy
wifi.rf.table <- wifi.rf$confusion[, -3]
sum(diag(wifi.rf.table)) / sum(wifi.rf.table)

plot(wifi.rf)
```

## Discussion

The confusion matrix output by `randomForest` is "the confusion matrix of the prediction (based on OOB data)".


# Problem 3 (10 points): Comparison to cross-validation tuned KNN predictor

Use convenience wrapper `tune.knn` provided by the library `e1071` on the entire dataset to determine optimal value for the number of the nearest neighbors 'k' to be used in KNN classifier.  Consider our observations from week 9 problem set when choosing range of values of `k` to be evaluated by `tune.knn`.  Setup resampling procedure similar to that used above for support vector classifier that will repeatedly: a) split WiFi localization dataset into training and test, b) use `tune.knn` on training data to determine optimal `k`, and c) use `k` estimated by `tune.knn` to make KNN classifications on test data.  Report and discuss distributions of test errors from this procedure and selected values of `k`, compare them to those obtained for random forest and support vector classifier above.

## Code

```{r tune.knn, cache=TRUE}
knn.acc <- numeric()
kvals <- numeric()
for (i in 1:N) {
  train <- sample(c(TRUE, FALSE), size = nrow(wifiLocDat), replace = TRUE)
  
  x <- subset(wifiLocDat[train, ], select = -Loc3)
  y <- wifiLocDat[train, ]$Loc3
  # Use 5-fold cross-validation for model
  res <- tune.knn(x, y, k = unique(round(2 ^ (seq(0, 7, by = 0.5)))),
                  tunecontrol = tune.control(sampling = "cross", cross = 5))
  
  p <- assess.prediction(wifiLocDat[!train,]$Loc3, 
                         knn(train = x, 
                             test = subset(wifiLocDat[!train, ], select = -Loc3),
                             cl = y,
                             k = res$best.parameters$k),
                         print.vals = FALSE)
  knn.acc <- c(knn.acc, sum(diag(p)) / sum(p))
  kvals <- c(kvals, res$best.parameters$k)
}
```

```{r tune.knn plots and analysis}
temp <- data.frame(measure = c(rep("accuracy", length(knn.acc)),
                               rep("k", length(kvals))),
                   value = c(knn.acc, kvals))
ggplot(data = temp, aes(x = "parameter", y = value, fill = measure)) + 
  facet_wrap(temp$measure, scales = "free") + 
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_jitter(height = 0)

plot(res)
plot(kvals, knn.acc, log = "x")

describe(knn.acc)
describe(kvals)
```

## Discussion

# Problem 4 (20 points): SVM with radial kernel

## Sub-problem 4a (10 points): impact of $gamma$ on classification surface

*Plot* SVM model fit to the WiFi localization dataset using (for the ease of plotting) *only the first and the second attributes* as predictor variables, `kernel="radial"`, `cost=10` and `gamma=5` (see ISLR Ch.9.6.2 for an example of that done with a simulated dataset).  You should be able to see in the resulting plot the magenta-cyan (or, in more recent versions of `e1071` -- yellow-brown) classification boundary as computed by this model.  Produce the same kinds of plots using 0.5 and 50 as values of `gamma` also.  Compare classification boundaries between these three plots and describe how they are impacted by the change in the value of `gamma`.  Can you trace it back to the role of `gamma` in the equation introducing it with the radial kernel in ISLR?

### Code

Adding in `gamma` of $0.1$, as default is `1/(data dimension)` per the documentation

```{r SVM gamma, cache=TRUE}
for (g in c(0.1, 0.5, 5, 50)){
  svm.result <-
    svm(Loc3 ~ .,
        data = wifiLocDat,
        kernel = "radial",
        cost = 10,
        gamma = g)
  # summary(svm.result)
  print(paste0("gamma = ", g, ":"))
  assess.prediction(actual = wifiLocDat$Loc3, 
                    predicted = predict(svm.result))
  plot(
    svm.result,
    data = wifiLocDat,
    formula = WiFi2 ~ WiFi1
  )
  title(sub = paste("gamma =", g))
}
```


## Sub-problem 4b (10 points): test error for SVM with radial kernel

Similar to how it was done above for support vector classifier (and KNN), set up a resampling process that will repeatedly: a) split the entire dataset (using all attributes as predictors) into training and test datasets, b) use `tune` function to determine optimal values of `cost` and `gamma` and c) calculate test error using these values of `cost` and `gamma`.  Consider what you have learned above about the effects of the parameters `cost` and `gamma` to decide on the starting ranges of their values to be evaluated by `tune`. Additionally, experiment with different sets of their values and discuss in your solution the results of it and how you would go about selecting those ranges starting from scratch.  Present resulting test error graphically, compare it to that of support vector classifier (with linear kernel), random forest and KNN classifiers obtained above and discuss results of these comparisons. 

### Code

I initially ran a small number of replicates ($n = 10$) with `cost = 2 ^ (-3:8)` and `gamma = sort(c(10 ^ (-3:3) / 2, 10 ^ (-3:3)))`. Since this is so computationally expensive, this let me narrow the useful range of both parameters:

|param|min|med|max|
|-----|---|---|---|
|cost |2  |12 |256|
|gamma|0.01|0.08|0.5|

A slightly larger number of replicates ($n = 25$) with `cost = 2 ^ (1:8)`, `gamma = sort(c(10 ^ (-2:0) / 2, 10 ^ (-2:0)))` then provided the results below, which let me narrow things further:

|param|min|med|max|
|-----|---|---|---|
|cost |2  |16 |256|
|gamma|0.01|0.1|0.5|

Final parameter range choice is `cost = 2 ^ (1:6)`, `gamma = c(0.05, 0.1, 0.25, 0.5)`

                            
```{r SVM gamma and cost, cache=TRUE}
svm.g.acc <- numeric()
costs.g <- numeric()
gammas <- numeric()
for (i in 1:N) {
  train <- sample(c(TRUE, FALSE), size = nrow(wifiLocDat), replace = TRUE)
  res <- tune(svm, Loc3 ~ ., data = wifiLocDat[train,],
              ranges = list(cost = 2 ^ (1:6),
                            gamma = c(0.05, 0.1, 0.25, 0.5)
                            )
              )
  p <- assess.prediction(wifiLocDat[!train,]$Loc3, 
                         predict(res$best.model, newdata = wifiLocDat[!train,]),
                         print.vals = FALSE)
  svm.g.acc <- c(svm.g.acc, sum(diag(p)) / sum(p))
  costs.g <- c(costs.g, res$best.parameters$cost)
  gammas <- c(gammas, res$best.parameters$gamma)
}
```

```{r SVM gamma plotting and analysis}
temp <- data.frame(measure = c(rep("accuracy", length(svm.g.acc)),
                               rep("cost", length(costs.g)), 
                               rep("gamma", length(gammas))), 
                   value = c(svm.g.acc, costs.g, gammas))
ggplot(data = temp, aes(x = "parameter", y = value, fill = measure)) + 
  facet_wrap(temp$measure, scales = "free") + 
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_jitter(height = 0)

plot(gammas, costs.g, log = "xy")
scatterplot3d(costs.g, gammas, svm.g.acc, type = "h", pch = 16, log = "x")

describe(svm.g.acc)
describe(costs.g)
describe(gammas)
```

### Discussion

TODO

# Extra 5 points problem: SVM with polynomial kernel

Repeat what was done above (plots of decision boundaries for various interesting values of tuning parameters and test error for their best values estimated from training data) using `kernel="polynomial"`.   Determine ranges of `coef0`, `degree`, `cost` and `gamma` to be evaluated by `tune`.  Present and discuss resulting test error and how it compares to linear and radial kernels and those of random forest and KNN.

## Code

A minimal set of `cost`, `degree`, and `coef0` used to avoid a combinatorial explosion of data (currently testing $3 \times 4 \times 4 = 48$ combinations)

```{r SVM polynomial, cache=TRUE}
svm.poly.acc <- numeric()
costs.p <- numeric()
degrees <- numeric()
coefs <- numeric()
for (i in 1:N) {
  train <- sample(c(TRUE, FALSE), size = nrow(wifiLocDat), replace = TRUE)
  res <- tune(svm, Loc3 ~ ., data = wifiLocDat[train,], kernel="polynomial",
              ranges = list(cost = c(1, 5, 10),
                            degree = 2:5,
                            coef0 = c(0, 0.1, 1, 10)
                            )
              )
  p <- assess.prediction(wifiLocDat[!train,]$Loc3, 
                         predict(res$best.model, newdata = wifiLocDat[!train,]),
                         print.vals = FALSE)
  svm.poly.acc <- c(svm.poly.acc, sum(diag(p)) / sum(p))
  costs.p <- c(costs.p, res$best.parameters$cost)
  degrees <- c(degrees, res$best.parameters$degree)
  coefs <- c(coefs, res$best.parameters$coef0)
}

temp <- data.frame(measure = c(rep("accuracy", length(svm.poly.acc)),
                               rep("cost", length(costs.p)), 
                               rep("degree", length(degrees)), 
                               rep("coef0", length(coefs))), 
                   value = c(svm.poly.acc, costs.p, degrees, coefs))
ggplot(data = temp, aes(x = "parameter", y = value, fill = measure)) + 
  facet_wrap(temp$measure, scales = "free") + 
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_jitter(height = 0)

describe(svm.poly.acc)
describe(costs.p)
describe(degrees)
describe(coefs)
```

``` {r}
for (deg in c(3, 4)){
  for (coef0 in c(0, 0.1, 1)){
    svm.result <-
      svm(Loc3 ~ .,
          data = wifiLocDat,
          kernel = "polynomial",
          cost = 1,
          degree = deg,
          coef0 = coef0
          )
    # summary(svm.result)
    print(paste0("cost = 1, degree = ", deg, ", coef0 = ", coef0, ":"))
    assess.prediction(actual = wifiLocDat$Loc3, 
                      predicted = predict(svm.result))
    plot(
      svm.result,
      data = wifiLocDat,
      formula = WiFi2 ~ WiFi1
    )
    title(sub = paste("Degree =", deg, ", coef0 =", coef0))
  }
}
```

## Discussion


